---
title: "Available datasets in opencodecounts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{available-datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Example

```{r setup}
# Load opencodecounts package
library(opencodecounts)
```

### SNOMED CT code usage in primary care in England

This is only a selection of the full dataset published by NHS Digital, for the data pre-processing see `/data-raw/snomed_code_usage.R`.

```{r snomed-usage}
# Return SNOMED code usage data
snomed_usage
```

### ICD-10 code usage in secondary care in England

This is the total annual count of the Finished Consultant Episodes (FCE) listing each 4-character ICD-10 code either in primary or secondary diagnosis position in the Hospital Episode Statistics in England.

This is only a selection of the full dataset published by NHS Digital, for the data pre-processing see `/data-raw/icd10_usage.R`.

```{r icd10-usage}
# Return ICD-10 code usage data
icd10_usage
```

### OPCS-4 code Usage in secondary care in England

This is the total annual count of each instance that each 4-character OPCS-4 code is listed across all primary and secondary procedure positions in the Finished Consultant Episodes (FCE) of the Hospital Episode Statistics in England.

This is only a selection of the full dataset published by NHS Digital, for the data pre-processing see `/data-raw/opcs4_usage.R`.

```{r opcs-4-usage}
# Return OPCS-4 code usage data
opcs4_usage
```