---
title: "How signposting coding has increased in general practice"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rise-of-signposting-codes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

"Signposting" describes directing a patient to a source of help (e.g., a pharmacist or a community service) without a formal referral.
In this example we searched all SNOMED CT descriptions for the terms "signpost" or "signposted" to see how often this activity has been recorded in general practice in England.

### Setup

```{r load-packages, message=FALSE, warning=FALSE}
# Load packages
library(opencodecounts)
library(dplyr)
library(stringr)
library(forcats)
library(ggplot2)
library(scales)
library(gt)
library(here)
```

```{r get-data}
# Search all GP SNOMED CT descriptions for "signpost" or "signposted"
signposting_usage <- get_gp_snomed() |>
  filter(str_detect(str_to_lower(description), "signpost")) |>
  mutate(
    description_short = strip_semantic_tag(description),
    description_short = str_replace(
      description_short,
      "Improving Access to Psychological Therapies",
      "IAPT"
    )
  )
```

`r n_distinct(signposting_usage$snomed_code)` distinct SNOMED CT codes have a description containing "signpost", e.g. "Signposting (procedure)", "Signposting to general practitioner (procedure)", and "Informal referral - signposted to other agency (finding)".

### Table with most used codes

```{r create-table}
# Calculate total usage by code for the entire time period
df_tab_top10_signposting <- signposting_usage |>
  group_by(snomed_code, description_short) |>
  summarise(total_usage = sum(usage), .groups = "drop") |>
  mutate(ratio_usage = total_usage / sum(total_usage)) |>
  slice_max(order_by = total_usage, n = 10)

# Create table
tab_top10_signposting <- df_tab_top10_signposting |>
  gt() |>
  cols_label(
    snomed_code = "SNOMED code",
    description_short = "Description",
    total_usage = "Usage",
    ratio_usage = "%"
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(font = "Courier New"),
    locations = cells_body(columns = snomed_code)
  ) |>
  cols_align(align = "left", columns = snomed_code) |>
  fmt_number(total_usage, decimals = 0) |>
  fmt_percent(ratio_usage, decimals = 1)

tab_top10_signposting

# Run this line manually to save the table as an image, e.g. to share on Slack
# gtsave(tab_top10_signposting, here("vignettes", "img", "tab_top10_signposting_total.png"))
```

The ten most used codes account for `r scales::percent(sum(df_tab_top10_signposting$total_usage) / sum(signposting_usage$usage), accuracy = 0.1)` of all recorded signposting activity.

### Figure showing change over time for the top 10 codes

```{r create-figure, fig.height=6, fig.width=7}
# Identify the 10 most used codes based on total usage
top10_signposting_codes <- df_tab_top10_signposting |>
  pull(snomed_code)

# Calculate yearly sum for the top 10 codes
df_fig_top10_signposting <- signposting_usage |>
  filter(snomed_code %in% top10_signposting_codes) |>
  group_by(end_date, description_short) |>
  summarise(usage = sum(usage), .groups = "drop") |>
  mutate(description_short = fct_reorder(
    description_short,
    usage,
    .fun = sum,
    .desc = TRUE
    )
  )

# x-axis breaks
x_breaks_signposting <- unique(df_fig_top10_signposting$end_date)

# Create figure
fig_top10_signposting <- df_fig_top10_signposting |>
  ggplot(
    aes(
      y = usage,
      x = end_date,
      colour = description_short
    )
  ) +
  geom_line(alpha = .7) +
  geom_point(size = 2) +
  labs(title = NULL, x = NULL, y = NULL, colour = NULL) +
  scale_x_date(breaks = x_breaks_signposting, labels = label_date("%b\n%Y")) +
  scale_y_continuous(labels = label_number(accuracy = 1), limits = c(0, NA)) +
  theme_classic() +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 5)) +
  scale_colour_brewer(palette = "Set3")

fig_top10_signposting

# Run this line manually to save the figure as an image, e.g. to share on Slack
# ggsave(here("vignettes", "img", "fig_top10_signposting_trends.png"), fig_top10_signposting, height = 6, width = 7)
```

```{r calc-yearly-trend, include = FALSE}
# Yearly totals, used for the inline numbers below
df_yearly_signposting <- signposting_usage |>
  group_by(end_date) |>
  summarise(total_usage = sum(usage), .groups = "drop") |>
  arrange(end_date)

date_first_year <- first(df_yearly_signposting$end_date)
date_last_year <- last(df_yearly_signposting$end_date)
usage_first_year <- first(df_yearly_signposting$total_usage)
usage_last_year <- last(df_yearly_signposting$total_usage)
usage_ratio_total <- usage_last_year / usage_first_year

# Calculate ratios
usage_2018 <- df_yearly_signposting |>
  filter(end_date == as.Date("2018-07-31")) |>
  pull(total_usage)

usage_2019 <- df_yearly_signposting |>
  filter(end_date == as.Date("2019-07-31")) |>
  pull(total_usage)

usage_2020 <- df_yearly_signposting |>
  filter(end_date == as.Date("2020-07-31")) |>
  pull(total_usage)

usage_2021 <- df_yearly_signposting |>
  filter(end_date == as.Date("2021-07-31")) |>
  pull(total_usage)

usage_ratio_2018_2019 <- usage_2019 / usage_2018
usage_ratio_2020_2021 <- usage_2021 / usage_2020
```

Recorded signposting activity increased from `r scales::comma(usage_first_year)` events in `r format(date_first_year, "%B %Y")` to `r scales::comma(usage_last_year)` by `r format(date_last_year, "%B %Y")`, around `r round(usage_ratio_total)` times as many.
Two periods stand out in the figure above: between July 2018 and July 2019, recording rose from `r scales::comma(usage_2018)` to `r scales::comma(usage_2019)` (`r round(usage_ratio_2018_2019, 1)` times as many), and between July 2020 and July 2021, from `r scales::comma(usage_2020)` to `r scales::comma(usage_2021)` (`r round(usage_ratio_2020_2021, 1)` times as many).
